|

### 2021-08-23 dissertation/msc_dissertation.Rmd
Final commit

---
title: |
  | __How analysis strategy affects analysis results__
subtitle: |
  | *Assessing results space and structure of @silberzahn2018 through model specification*
  |
  | MSc Psychological Research Methods with Data Science
  | Dissertation
author: "Sebastian Ploner, sploner1@sheffield.ac.uk, University of Sheffield"
output:
  pdf_document:
    includes:
      in_header: header.tex
    fig_caption: yes
  word_document: default
linestretch: 1.5
urlcolor: blue
bibliography: references.bib
csl: apa.csl
header-includes:
- \usepackage[ruled]{algorithm2e}
- \usepackage{float}
- \usepackage{caption}
- \captionsetup[figure]{font={stretch=1}}
---

@@

# Abstract
Crowdsourcing and multiverse analysis usually only assess reasonable analysis strategies. Little is known about how the results space of these *reasonable* analyses relates to the results space of all *possible* analyses. The current project, therefore, aimed at defining and understanding the results space and structure, respectively, of all possible analyses for @silberzahn2018. Due to time and resources limitations, the project focused on the analysis producing the median outcome and randomly sampled 1,000 covariate combinations out of all possible ones. The findings indicate that the full results space is narrower than the original confidence intervals suggest. If this were to be true for all analyses conducted in @silberzahn2018, the variability between the analysis strategies were likely smaller than previously though. Future studies have to confirm this hypothesis by first running the "median model" with all covariate combinations and for each statistical model.

@@

 Another approach to assess the effects of analysis strategies on research findings is multiverse analysis [also known as specification-curve analysis, @simonsohn2020]. This approach has been proposed by @botvinik-nezer2020 and @silberzahn2018 as an alternative to crowdsourcing. In this approach, all reasonable analysis strategies are identified and performed. *Reasonable strategies* are defined as all statistically valid, non-redundant tests that are appropriate for the research question. The aggregated results of all reasonable analysis strategies are used to make inferences about the research question. Although the approach is statistically complex and computationally intense, it can be performed by one or a few researchers. Moreover, a researcher’s inherent strategy bias is neutralised, and noise is made transparent [@simonsohn2020]. Multiverse approaches are therefore well suited to better understand how different analysis strategies influence research results.

@@

# Analysis
## Analysis plan
     The project's objective was to define the space and the structure of all possible results of @silberzahn2018. The results space refers to the numerical interval between the lowest and highest possible outcome. It is created by running every possible analysis strategy. Hence, it is closely related to assessing the robustness of an effect. Robustness refers to the consistency of an effect under different model specifications. @patel2015 developed a standardised approach to assess an effect's robustness. This standardised approach to define the results space was used. The approach essentially comprises two parameters: the statistical models and the covariates (or control variables). In @silberzahn2018, the analysts used numerous statistical models like multiple linear regression, mixed-model logistic regression or Bayesian logistic regression. In total, there were 29 different modelling approaches. Additionally, each team used a different set of covariates. Across all teams, 15 covariates were used, resulting in $2^n$, i.e., $2^{15} = 32,768$ possible combinations of covariates. Adding all modelling possibilities to the equation yields a total of $29 * 2^{15} = 950,272$ possible analysis strategies. As this number of strategies exceeds the current project's resources, it was reduced to a more manageable number. Therefore, the project focused on the modelling approach producing the median outcome of all analyses. The median, being the middle number of a given set of values, is a reasonable starting point to estimate the results space. However, even focusing on one analysis approach leaves $1 * 2^{15} = 32,768$ possibilities. Due to computational limitations, a random sample of 1,000 analysis strategies were conducted.

 In @silberzahn2018, the median outcome was produced by @team23, which will hereinafter be referred to as *Team 23* due to its team number in the original study. *Team 23* first transformed the data and then conducted a mixed-model logistic regression. The first step of the current project was to replicate their transformation and analysis. Such replication can increase the confidence in the previous and the current approach. Moreover, it ensures that this project has the same starting point as *Team 23*. As team made all scripts publicly available, this step was straightforward. The next step of the project was to define the results space and its structure. To this end, a random sample of covariates without replacement was drawn. “Without replacement” ensured that all covariate combinations were unique in the sample. The covariates were appended to the base (or core) variables. Base variables were those variables that were primarily assessed to answer the research question. In this case, the research question was whether a football player’s skin colour affects the odds of being sent off the field. *Team 23* defined two interaction terms as the base: “skin tone X implicit bias” and “skin tone X explicit bias.” (The variables are described in the “data” section.) Hence, all models had the following structure:
$$
\begin{aligned}
RedCard = SkinTone*ImplicitBias + SkinTone*ExplicitBias + CovariateCombination_{i}
\end{aligned}
$$ 
The relevant outcome parameters of each model were extracted. These parameters were the coefficient (i.e., effect) of skin tone, its standard error, test statistic and p-value. Similar to *Team 23* the 95% confidence intervals (CI) were calculated. The estimates and their CIs were transformed to odds ratios (OR) through exponentiating them to the power of two. OR quantify the strength of association between two variables. An OR greater than one indicates that the dependent variable is more likely to occur given the independent variable, if it is lower than one it is less likely to occur. Eight OR outliers were excluded from visualisation. Four approached infinity, one was in the lower million range, one was 150, and a final one was 4.6.

@@

## Exploring the results space - specification curve
     Second, the specification curve was calculated to visualise the results space (Figure \ref{fig:sca_plot}). The horizontal axes of the top and bottom chart show the specifications sorted by their OR from lowest to highest. The OR are shown on the vertical axis of the top plot. The points in the top plot (which together look like a line) each represent the outcome of one statistical model. Black and red points refer to statistically significant and non-significant outcomes, respectively. Overall, 89.7% of the outcomes were significant. Except for the lowest outcome ($OR_{lowest} = 1.081$) all points were close to their neighbours. This indicates that none of the specified models was responsible for a sudden increase in the outcome measure. The bottom table lists all covariates on its vertical axis, sorted from most positive (top) to most negative impact (bottom). For instance, the covariate "player" had the most substantial positive impact, while the covariate "club" had the most substantial negative impact. Each column of the table represents one model. A coloured cell indicates the presence of the variable, an uncoloured cell its absence. Again, black and red colours indicate that the outcome measure was significant or non-significant, respectively.

@@

## Exploring the results space – raincloud plot
     Third, the results space was further explored by calculating raincloud plots (Figure \ref{fig:results_space_plot}). The horizontal axis shows the OR. From top to bottom, the three components show the probability density distribution, the box plot and the raw data. Based on the raw data points, the results space can be defined as the interval between 1.081 and 1.383. The 1st quartile was 1.206, the median 1.248 and the 3rd quartile 1.293. Thus, the interquartile range where 50% of the data were included was relatively narrow (1.206 to 1.293), indicating a low statistical dispersion. The dashed line indicates the original result of *Team 23*, which has been the median outcome of all analysis strategies in @silberzahn2018. The outcome of *Team 23* (1.31) was included in the results space but outside the middle 50% of the data. The CI calculated by team 23 ranged from 1.10 to 1.56. The lower bound of their CI was close to the lower margin of the results space of the present study. In contrast, the upper bound of their CI was notably higher than the present results space. This suggests a potential overestimating of the variability of the results in the original analysis.

@@

```{r results_space_plot, fig.cap = "Raincloud plot. The results space ranges from about 1.08 to 1.38. The distribution showed two peaks which suggests an influential latent covariate structure. The original median effect was within the results space but outside the middle 50 percent of the data."}

@@

## Specifying covariate effects – scatter plot 
     Fourth, to better understand the relevance of the covariates, another scatter plot was calculated (Figure \ref{fig:covariate_effects_plot}). This graph shows the specified covariate effects. On the horizontal axis are the covariates; on the vertical axis are their estimated effects. The covariates are sorted from negative to positive impact. Black indicates a significant effect, red non-significant. The error bars represent the 95% CI. Four out of 15 covariates were significant. All effect sizes, including the significant covariates’, were relatively weak. The CIs were relatively small, which was likely due to the large sample size. The OR of the model without any covariates (i.e., core variables only) was 1.262. Adding the sum of all positive covariate effects (games + goals + referee country + player, respectively) to the base model resulted in a maximum possible effect of 1.357. The maximum observed OR was 1.382, however. This difference suggests that there were more impactful covariate combinations or interactions that have not yet been identified. Vice versa, subtracting the sum of all negative covariate effects from the base model resulted in an OR of 1.070, whereas the minimum observed effect was an OR of 1.081. This suggests that there were also covariate combinations with a higher negative impact. \newline

@@

## Exploring the results structure – volcano plot
     Lastly, another graph was produced despite not being planned. Therefore, it is only used to strengthen previous points and probe future research. The results above suggest one or more meaningful latent covariate combinations; these were further explored. Although @patel2015 had a different objective, their methodology was similar to the current project. They visualised the outcomes as a so-called volcano plot (see Figure \ref{fig:latent_structure_plot}), showing the relationship between the p-values and effect sizes. A similar graph was calculated for the current project. The two (potentially three) distinct “streams” provided further support for two (or three) distinct underlying constructs. Future research might specify which covariate combinations underly these structures.

@@

 Moreover, the distribution of the results space showed two peaks. To understand the distribution’s underlying structure, the covariate effects were analysed. The results showed that 4 out of 15 covariates were statistically significant, though their effects were small. Moreover, when adding the individual covariate effects to the outcome of the base model, i.e. the model without any covariates, the observed outcome values are always well within the results space. Thus, interactions of covariates likely contribute to the results space. The final volcano plot supports this hypothesis by showing at least two streams indicating at least two underlying constructs. Future research might investigate these latent structures. In short, the current project found that the full results space is narrower than the CIs of the original “median” analysis of @silberzahn2018 indicated. This suggests that the original study overestimated the variability. This was surprising as the full results space would have been expected to be larger than the estimated CIs. If the variability were overestimated for all analyses, this would suggest that there is more agreement between the analyses than previously assumed.

>>>>

---
title: |
  | __How analysis strategy affects analysis results__
subtitle: |
  | *Assessing results space and structure of @silberzahn2018 through model specification*
  |
  | MSc Psychological Research Methods with Data Science
  | Dissertation
author: "Registration no.: 200256786"
output:
  pdf_document:
    includes:
      in_header: header.tex
    fig_caption: yes
  word_document: default
linestretch: 1.5
urlcolor: blue
bibliography: references.bib
csl: apa.csl
header-includes:
- \usepackage[ruled]{algorithm2e}
- \usepackage{float}
- \usepackage{caption}
- \captionsetup[figure]{font={stretch=1}}
---

@@

# Abstract

Most research studies can be analysed by using different strategies which can cause variability in the results. Crowdsourcing and multiverse analysis are recent approaches to better understand the influence of analysis strategies on research findings. These approaches usually assess different *reasonable* but not all *possible* analysis strategies. Thus, little is known about how the results space of these *reasonable* analyses relates to the results space of all *possible* analyses. The current project therefore aimed at defining and understanding the results space and structure of all possible analyses for an exemplary study by @silberzahn2018. To address this topic, the project focused on one statistical model producing the median outcome and calculated the model for 1,000 randomly sampled covariate combinations. The findings indicate that the full results space is narrower than the original confidence intervals suggest. Thus, the variability between the analysis strategies in @silberzahn2018 was likely smaller than previously thought. Future investigations might extend the present study by running even more statistical models with all possible covariate combinations.

@@

 Another approach to assess the effects of analysis strategies on research findings is multiverse analysis [also known as specification-curve analysis, @simonsohn2020]. This approach has been proposed by @botvinik-nezer2020 and @silberzahn2018 as an alternative to crowdsourcing. In this approach, all reasonable analysis strategies are identified and performed. *Reasonable strategies* are defined as all statistically valid, non-redundant tests that are appropriate for the research question. The aggregated results of all reasonable analysis strategies are used to make inferences about the research question. Although the approach is statistically complex and computationally intense, it can be performed by one or a few researchers. Moreover, it neutralises a researcher’s inherent strategy bias and makes noise transparent [@simonsohn2020]. Multiverse approaches are therefore well suited to better understand how different analysis strategies influence research results.

@@

# Analysis
## Analysis plan
     The project's objective was to define the space and the structure of all possible results of @silberzahn2018. The results space refers to the numerical interval between the lowest and highest possible outcome. It is created by running every possible analysis strategy. Hence, it is closely related to assessing the robustness of an effect. Robustness refers to the consistency of an effect under different model specifications. @patel2015 developed a standardised approach to assess an effect's robustness. This standardised approach to define the results space was used. The approach essentially comprises two parameters: the statistical models and the covariates (or control variables). In @silberzahn2018, the analysts used numerous statistical models like multiple linear regression, mixed-model logistic regression or Bayesian logistic regression. In total, they used 29 different modelling approaches. Additionally, each team used a different set of covariates. Across all teams, 15 covariates were used, resulting in $2^n$, i.e., $2^{15} = 32,768$ possible combinations of covariates. Adding all modelling possibilities to the equation yields a total of $29 * 2^{15} = 950,272$ possible analysis strategies. As this number of strategies exceeds the current project's resources, it was reduced to a more manageable number. Therefore, the project focused on the modelling approach producing the median outcome of all analyses. The median is the middle number of a given set of values and therefore a reasonable starting point to estimate the results space. However, even focusing on one analysis approach leaves $1 * 2^{15} = 32,768$ possibilities. Due to computational limitations, a random sample of 1,000 analysis strategies were conducted.

 In @silberzahn2018, the median outcome was produced by @team23, which will hereinafter be referred to as *Team 23* due to its team number in the original study. *Team 23* first transformed the data and then conducted a mixed-model logistic regression. The first step of the current project was to replicate their transformation and analysis. Such replication can increase the confidence in the previous and the current approach. Moreover, it ensures that this project has the same starting point as *Team 23*. As *Team 23* made all scripts publicly available, this step was straightforward. The next step of the project was to define the results space and its structure. To this end, a random sample of covariates without replacement was drawn. “Without replacement” ensured that all covariate combinations were unique in the sample. The covariates were appended to the base (or core) variables. Base variables were those variables that were primarily assessed to answer the research question. In this case, the research question was whether a football player’s skin colour affects the odds of being sent off the field. *Team 23* defined two interaction terms as the base: “skin tone X implicit bias” and “skin tone X explicit bias.” (The variables are described in the “data” section.) Hence, all models had the following structure:
$$
\begin{aligned}
RedCard = SkinTone*ImplicitBias + SkinTone*ExplicitBias + CovariateCombination_{i}
\end{aligned}
$$ 
The relevant outcome parameters of each model were extracted. These parameters were the coefficient (i.e., effect) of skin tone, its standard error, test statistic and p-value. Similar to *Team 23* the 95% confidence intervals (CI) were calculated. The estimates and their CIs were transformed to odds ratios (OR) through exponentiating them to the power of two. OR quantify the strength of association between two variables. An OR greater than one indicates that the dependent variable is more likely to occur given the independent variable, if it is lower than one it is less likely to occur. Eight OR outliers were excluded from visualisation. Four approached infinity, one was in the lower million range, one was 150, and a final one was 4.6.

@@

## Exploring the results space - specification curve
     Second, the specification curve was calculated to visualise the results space (Figure \ref{fig:sca_plot}). The horizontal axes of the top and bottom chart show the specifications sorted by their OR from lowest to highest. The OR are shown on the vertical axis of the top plot. The points in the top plot (which together look like a line) each represent the outcome of one statistical model. Black and red points refer to statistically significant and non-significant outcomes, respectively. Overall, 89.7% of the outcomes were significant. Except for the lowest outcome ($OR_{lowest} = 1.081$) all points were close to their neighbours. This indicates that none of the specified models caused a sudden increase in the outcome measure. The bottom table lists all covariates on its vertical axis, sorted from most positive (top) to most negative impact (bottom). For instance, the covariate "player" had the most substantial positive impact, while the covariate "club" had the most substantial negative impact. Each column of the table represents one model. A coloured cell indicates the presence of the variable, an uncoloured cell its absence. Again, black and red colours indicate that the outcome measure was significant or non-significant, respectively.

@@

## Exploring the results space – raincloud plot
     Third, the results space was further explored by calculating raincloud plots (Figure \ref{fig:results_space_plot}). The horizontal axis shows the OR. From top to bottom, the three components show the probability density distribution, the box plot and the raw data. Based on the raw data points, the results space can be defined as the interval between 1.081 and 1.383. The 1st quartile was 1.206, the median 1.248 and the 3rd quartile 1.293. Thus, the interquartile range where 50% of the data were included was relatively narrow (1.206 to 1.293), indicating a low statistical dispersion. The dashed line indicates the original result of *Team 23*, which has been the median outcome of all analysis strategies in @silberzahn2018. The outcome of *Team 23* (1.31) was included in the results space but outside the interquartile range. The CI calculated by team 23 ranged from 1.10 to 1.56. The lower bound of their CI was close to the lower margin of the results space of the present study. In contrast, the upper bound of their CI was notably higher than the present results space. This suggests a potential overestimation of the variability of the results in the original analysis.

@@

```{r results_space_plot, fig.cap = "Raincloud plot. The results space ranges from about 1.08 to 1.38. The distribution showed two peaks which suggests an influential latent covariate structure. The original median effect was within the results space but outside the interquartile range."}

@@

## Specifying covariate effects – scatter plot 
     Fourth, to better understand the relevance of the covariates, another scatter plot was calculated (Figure \ref{fig:covariate_effects_plot}). This graph shows the specified covariate effects. On the horizontal axis are the covariates; on the vertical axis are their estimated effects. The covariates are sorted from negative to positive impact. Black indicates a significant effect, red non-significant. The error bars represent the 95% CI. Four out of 15 covariates were significant. All effect sizes, including the significant covariates' ones, were relatively weak. The CIs were relatively small, which was likely due to the large sample size. The OR of the model without any covariates (i.e., core variables only) was 1.262. Adding the sum of all positive covariate effects (games + goals + referee country + player, respectively) to the base model resulted in a maximum possible effect of 1.357. The maximum observed OR was 1.382, however. This difference suggests that there were more impactful covariate combinations or interactions that have not yet been identified. Vice versa, subtracting the sum of all negative covariate effects from the base model resulted in an OR of 1.070, whereas the minimum observed effect was an OR of 1.081. This suggests that there were also covariate combinations with a higher negative impact. \newline

@@

## Exploring the results structure – volcano plot
     Lastly, another graph, which had not been planned, was produced. Therefore, it is only used to strengthen previous points and probe future research. The results above suggest one or more meaningful latent covariate combinations; these were further explored. Although @patel2015 had a different objective, their methodology was similar to the current project. They visualised the outcomes as a so-called volcano plot (see Figure \ref{fig:latent_structure_plot}), showing the relationship between the p-values and effect sizes. A similar graph was calculated for the current project. The two (potentially three) distinct “streams” provided further support for two (or three) distinct underlying constructs. Future research might specify which covariate combinations underly these structures.

@@

 Moreover, the distribution of the results space showed two peaks. To understand the distribution’s underlying structure, the covariate effects were analysed. The results showed that 4 out of 15 covariates were statistically significant, though their effects were small. Moreover, when adding the individual covariate effects to the outcome of the base model, i.e. the model without any covariates, the observed outcome values are always well within the results space. Thus, interactions of covariates likely contribute to the results space. The final volcano plot supports this hypothesis by showing at least two streams indicating at least two underlying constructs. Future research might investigate these latent structures. Moreover, the current project found that the full results space is narrower than the CIs of the original “median” analysis of @silberzahn2018 indicated. This suggests that the original study overestimated the variability. This was surprising as the full results space would have been expected to be larger than the estimated CIs. If the variability were overestimated for all analyses, this would suggest that there is more agreement between the analyses than previously assumed.
